---
title: "Daff"
author: "Edwin de Jonge (@edwindjonge), Gregory Warnes"
date: "UseR!2017, July 6, 2017"
output:
  beamer_presentation:
    keep_tex: yes
    includes:
      in_header: header.tex
    theme: "Frankfurt"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(magrittr)
```

## Who am I (Edwin)

- Statistical consultant / methodologist
- Statistics Netherlands (CBS): all official statistics of NL
- US Census + Labor Statistics + Economic Statistics + Agriculture etc.

## What is daff?

### Short version

> Daff is a diff for `data.frame`s

- Detect changes: **`diff_data`**, **`differs_from`**
- Store and restore diff: **`write_diff`**, **`read_diff`**
- Patch updated data: **`patch_data`**, **`merge_data`**
- Render a diff: **`render_diff`**

## Why o why?
- log changes of data.
- support version control of `data.frame`
- subsequent steps in data process: what did these steps do?
- Supports version control of data. 
- Just like `diff` for source code, but optimized for `data.frame`.

## Use case: data update

### Raw data update

- You have build a nice R script:
    - takes raw data as input
    - removes errors
    - fits a model
    - calculates output

You get an updated raw data file: what are the changes?

## Use case: manual editing

- out of your control, there is a manual editing step:
- *bad practice*, but it happens: e.g. implausible values, manual data correction.

### Manual editing

\begin{center}
  \includegraphics[height=0.3\textheight]{fig/blackbox}
\end{center}
- Compare the input and output
- Make the manual step _reproducible_: all process steps can be automated.

## Detecting changes

`daff` detects the following changes:

- changing a value
- adding a row
- removing a row
- adding a column
- removing a column
- changing type of a column (partially)

## `diff_data`: value change

```{r, echo=TRUE}
library(daff)
x         <- data.frame(a=1, b=2)
x_changed <- data.frame(a=1, b=3)
p <- diff_data(x, x_changed)
print(p)
```
## `patch_data`: apply the change

```{r, echo=TRUE}
x
patch_data(x, p)
```


## detect a row addition

```{r, echo=TRUE}
x <- data.frame(a=1, b=2)
x_changed <- data.frame(a=1:2, b=2:3)
diff_data(x,x_changed)
```

## `diff_data`: detect a row deletion

```{r, echo=TRUE}
x <- data.frame(a=1:2, b=2:3)
x_changed <- data.frame(a=1, b=2)
diff_data(x,x_changed)
```

## diff_data options

```{r, echo=TRUE, eval=FALSE}
diff_data( data_ref, data
         , always_show_header       = TRUE
         , always_show_order        = FALSE
         , columns_to_ignore        = c()
         , count_like_a_spreadsheet = TRUE
         , ids                      = c()
         , ignore_whitespace        = FALSE
         , never_show_order         = FALSE
         , ordered                  = TRUE
         , ... 
         )
```


## Implementation

- Wraps the excellent library `daff` javascript, by Paul Fitzpatrick (@fitzyfitzyfitzy).
- library actually written in Haxe, which compiles to js, python, C++
- Uses R package `V8` to run javascript (Thanks Jeroen Ooms!)

## 

\Large{Thank you for your attention!}

Interested?

```{r, echo=TRUE, eval=FALSE}
install.packages("daff")`
```

or visit: 

- http://github.com/edwindj/daff
